# 研究設計入門 {#sec-A-brief-introduction-to-research-design}

```{r}
#| include: FALSE
source("header.R")
```

> *"實驗完成後才諮詢統計學家，只是請他幫忙驗屍而已。統計學家只能為你開出實驗的死亡證明。"*
> -- 羅納德·愛爾默·費雪爵士[^02-a-brief-introduction-to-research-design-1]

[^02-a-brief-introduction-to-research-design-1]: 節錄自1938年，第一屆印度統計學會主席演講。來源：[http://en.wikiquote.org/wiki/Ronald Fisher](http://en.wikiquote.org/wiki/Ronald Fisher)

<!---
> *"To consult the statistician after an experiment is finished is often merely to ask him to conduct a post mortem examination. He can perhaps say what the experiment died of."*\
> -- Sir Ronald Fisher[^02-a-brief-introduction-to-research-design-1]

[^02-a-brief-introduction-to-research-design-1]: Presidential Address to the First Indian Statistical Congress, 1938. Source: [http://en.wikiquote.org/wiki/Ronald Fisher](http://en.wikiquote.org/wik%0Ai/RonaldFisher) --->


這一章是談入門研究設計要知道的基本概念，像是收集資料，如何確認收集的資料是符合目的等等。雖然同學們讀完這一章無法開始一項真正的研究，卻能掌握解讀與評估已公開發表研究設計的初級工具。由於這本書的學習重點是資料分析，而不是如何收集資料，這一章只會給你很簡單的介紹。這一章有兩個“特別之處”，首先這一章談到的心理學知識比之後的章節多，其次這一章討論研究方法的科學問題比資料分析的統計問題還多。科學問題與統計問題密不可分，傳統統計書考慮篇幅，比較少討論科學問題。這一章大部分內容來自 @Campbell1963 與 @Stevens1946 關於測量尺度的討論。

<!---In this chapter, we're going to start thinking about the basic ideas that go into designing a study, collecting data, checking whether your data collection works, and so on. It won't give you enough information to allow you to design studies of your own, but it will give you a lot of the basic tools that you need to assess the studies done by other people. However, since the focus of this book is much more on data analysis than on data collection, I'm only giving a very brief overview. Note that this chapter is "special" in two ways. Firstly, it's much more psychology specific than the later chapters. Secondly, it focuses much more heavily on the scientific problem of research methodology, and much less on the statistical problem of data analysis. Nevertheless, the two problems are related to one another, so it's traditional for stats textbooks to discuss the problem in a little detail. This chapter relies heavily on @Campbell1963 and @Stevens1946 for the discussion of scales of measurement. --->

## 認識心理測量 {#sec-Introduction-to-psychological-measurement}

首先請同學理解資料分析就是一種**測量**。在心理學系要學的就是如何測量人類行為及心智活動。那麼到底"測量"是什麼？

<!---The first thing to understand is data collection can be thought of as a kind of **measurement**. That is, what we're trying to do here is measure something about human behaviour or the human mind. What do I mean by "measurement"?--->

### 心理測量面面觀

測量並不是一個實在的概念，而是泛指賦予可觀察的屬性數值、標記、或者某種清晰定義之類的表述。以下都是心理測量的實際例子：

- 我的**年齡**是*33*歲。
- *我不喜歡* **小魚乾**。
- 我的**染色體性別**是*男性*。
- 我的**自我認同性別**是*女性*。

以上例子裡的**粗體字**是“可被測量的事物或 屬性”，*斜體字*就是“測量值”[^02-translation-1]。我們可以再對每個例子做些深入探討，了解每種測量的細節：

[^02-translation-1]: 譯註~英文版提到“測量程序”與"測量值“都是以”measurement"稱呼，中文版根據主題及脈絡以不同名稱區別。

- 我的**年齡**(以年計)可以是*0,1,2,3...*。年齡的上限是多少沒人知道，不過你可以假定我能活到*150*歲，雖然還沒有人類活到這麼老。
- 若你問我**喜歡小魚乾**嗎？我可以回答*我喜歡*，*我不喜歡*，*我沒有喜不喜歡*，*我有時候喜歡*。
- 我的**染色體性別**在多數情況下是*男性($XY$)*或*女性($XX$)*，不過凡事總有例外。我可能有*克林那費爾特症候群($XXY$)*，使得我有男性和女性性徵。當然還有其他性染色體變異的可能。
- 我的**自我認同性別**也許傾向男性或女性，而這和我的染色體性別無關。我也可以認為自已不是二元性別，也可以宣稱是*跨性別*。

如你所見，有的例子像是能間單測量數值的項目，例如年齡，但是有些研究主題的特殊考量會讓測量變得複雜。因為年齡是大家都能理解的例子，這裡先用年齡做個說明。讀到這裡的同學可能會想，用“年度”做為測量年齡的數值不就好了嗎？若是你加入發展心理學的研究，“年度”可能不是精確的測量單位，可能還有加上“月份”，例如找到兒童參與者2歲11個月大，可縮寫為"2:11"。如果研究對象是新生兒，可能要紀錄出生到現在的“天數”，甚至要用“小時”紀錄。總而言之，確認在研究主題之下，什麼是有意義的測量數值是重要的第一課。

再想得更仔細的話，我們會發覺“年齡”並不是全然精確的概念。在一般情況，我們講“年齡”有隱含“從出生到現在的時間長度”的意思。但是並不是每次提到“年齡”就是這樣的意思。設想在一個研究新生兒控制眼球運動的實驗室，受測幼兒的"出生時間”不一定是實驗人員記錄“年齡”的有意義參照點。如果今天找到兩位“出生兩小時”的新生兒，一位是比預產期提前三週出生的早產兒，另一位比預產期晚一週出生，那麼研究人員能紀錄這兩位新生兒“年齡相同”嗎？從日常社會互動的情境來說，每個人都是以出生時間做為計算年齡的參照點，因為任何人都能接受每個人出生後都是獨立生活的個體。然而在科學研究的情境要考慮的就不只如此。將人類當成一種生物的話，最能獲得充分分析的年齡紀錄，應該從個體受孕那一刻起，紀錄成長到成熟的狀況。所以“年齡”的測量定義有兩種：從受孕時間起算的年齡，以及從出生時間起算的年齡。研究對象是成人的話或許不必計較兩種定義的區別，若研究對象是新生兒也許要考量那種定義才符合研究目的。

由測量定義問題再討論下去，就是**方法學**的課題了。有關人類年齡的”測量方法“，同學們能想到或找到幾種有用的調查方法呢？以下是一些你可以參考的做法： 

- 直接問受測對象”你現在幾歲？“ 這種自陳式報告是快速、便宜又簡單的調查方法。不過必須確認受測對象已經成長到有能力回答調查問題，並且不會給不實回答。
- 詢問可靠的消息來源，像是受測幼童的父母親“你們的孩子幾歲？”這種調查方法實施快速，特別是受測者的父母親就在旁邊的情境。不過如果要調查的是"受孕年齡“，大多數父母親都不清楚是什麼時候開始懷孕的。這時也許要詢問更可靠的來源，像是他們的婦產科主治醫師。
- 尋找官方紀錄，像是個人的出生與死亡證明。這種方式耗時又勞力，不過如果研究對象是已往生的人是種可靠的來源。

<!---
Measurement itself is a subtle concept, but basically it comes down to finding some way of assigning numbers, or labels, or some other kind of well-defined descriptions, to "stuff". So, any of the following would count as a psychological measurement:

-   My **age** is 33 years.
-   I do not **like anchovies.**
-   My **chromosomal gender** is male.
-   My **self-identified gender** is female.

In the short list above, the **bolded part** is "the thing to be measured", and the *italicised part* is "the measurement itself". In fact, we can expand on this a little bit, by thinking about the set of possible measurements that could have arisen in each case:

-   My **age** (in years) could have been *0, 1, 2, 3* ..., etc. The upper bound on what my age could possibly be is a bit fuzzy, but in practice you'd be safe in saying that the largest possible age is *150*, since no human has ever lived that long.
-   When asked if I **like anchovies**, I might have said that *I do, or I do not, or I have no opinion, or I sometimes do.*
-   My **chromosomal gender** is almost certainly going to be *male (*$XY$) or *female (*$XX$), but there are a few other possibilities. I could also have *Klinfelter's syndrome* ($XXY$), which is more similar to male than to female. And I imagine there are other possibilities too.
-   My **self-identified** gender is also very likely to be male or female, but it doesn't have to agree with my chromosomal gender. I may also choose to identify with *neither*, or to explicitly call myself *transgender*. 

As you can see, for some things (like age) it seems fairly obvious what the set of possible measurements should be, whereas for other things it gets a bit tricky. But I want to point out that even in the case of someone's age it's much more subtle than this. For instance, in the example above I assumed that it was okay to measure age in years. But if you're a developmental psychologist, that's way too crude, and so you often measure age in *years and months* (if a child is 2 years and 11 months this is usually written as "2;11"). If you're interested in newborns you might want to measure age in *days since birth*, maybe even *hours since birth*. In other words, the way in which you specify the allowable measurement values is important.


Looking at this a bit more closely, you might also realise that the concept of "age" isn't actually all that precise. In general, when we say "age" we implicitly mean "the length of time since birth". But that's not always the right way to do it. Suppose you're interested in how newborn babies control their eye movements. If you're interested in kids that young, you might also start to worry that "birth" is not the only meaningful point in time to care about. If Baby Alice is born 3 weeks premature and Baby Bianca is born 1 week late, would it really make sense to say that they are the "same age" if we encountered them "2 hours after birth"? In one sense, yes. By social convention we use birth as our reference point for talking about age in everyday life, since it defines the amount of time the person has been operating as an independent entity in the world. But from a scientific perspective that's not the only thing we care about. When we think about the biology of human beings, it's often useful to think of ourselves as organisms that have been growing and maturing since conception, and from that perspective Alice and Bianca aren't the same age at all. So you might want to define the concept of "age" in two different ways: the length of time since conception and the length of time since birth. When dealing with adults it won't make much difference, but when dealing with newborns it might.

Moving beyond these issues, there's the question of methodology. What specific "measurement method" are you going to use to find out someone's age? As before, there are lots of different possibilities:

-   You could just ask people "how old are you?" The method of self-report is fast, cheap and easy. But it only works with people old enough to understand the question, and some people lie about their age.
-   You could ask an authority (e.g., a parent) "how old is your child?" This method is fast, and when dealing with kids it's not all that hard since the parent is almost always around. It doesn't work as well if you want to know "age since conception", since a lot of parents can't say for sure when conception took place. For that, you might need a different authority (e.g., an obstetrician).
-   You could look up official records, for example birth or death certificates. This is a time consuming and frustrating endeavour, but it has its uses (e.g., if the person is now dead).--->

### 操作型定義: 定義研究的測量方法

以上討論這麼多，是要引導同學理解**操作型定義**的概念。用更清楚簡潔的話來說，操作型定義是*將有意義但有些模糊的概念，轉換為精確測量程序的設定過程*。完整的操作型定義要在過程中考慮這些條件：

- 清楚定義研究要測量的項目。以“年齡”為例，研究要測量的是“胎兒出生起算的時間長度”還是“母體受孕起算的時間長度”。
- 確認用什麼方法進行測量。同樣以“年齡”為例，研穵 要用自陳式報告、詢問受測對象雙親、還是查閱官方紀錄？如果決定用自陳式報告，要如何設定問題內容？
- 定義可做紀錄的測量數值範圍。請留意數值不一定要是數字！雖然年齡的測量數值無疑問題數字，依然要考慮什麼單位的數字是需要被紀錄的。是年份？月份？還是小時？像性別一類數值不是數字的測量方法，同樣也要考慮數值的範圍。像是請受測對象透過自陳報告回報性別，要設計什麼樣的選項讓人選擇？只有“男性”和“女性”兩個選項夠嗎？需不需要增加“其他”這一項？或者改用開放式回答，讓受測對象自已寫下個人認同的性別？如果確定使用開放式回答，要如何解讀各式各樣的答案？

設定操作型定義有各種方法，沒有一種“唯一正解”。任何研究都是根據測量目的，透過設定操作型定義的過程，將“年齡”與"性別"等不具形式的概念，轉化為可用數值形式表達的測量值。每個科學領域都對研究對象的測量方式有基本共識。因此，要理解如何設定操作型定義，是因研究主題而異。畢竟有許多主題存在大量差異化的個別研究，有些主題則有一致的研究模式，無法用統一的方法設定操作型定義。

進入下一節之前，讓我們整理一些之後的單元會經常遇到，與測量有關的專用名詞，這些名詞彼此之間有許多關聯：

- **理論建構(A theoretical construct)** 研究者想要測量的目標，像是"年齡"、"性別"、或"選項"。理論建構不能直接觀察記錄，都是模糊抽象的概念。
- **測量程序(measure)[^02-transltion-02]** 測量程序是進行觀察紀錄的方法或工具。像是問卷的題目、行為觀察、腦部活動掃瞄都是一種測量程序。
- **操作型定義(operationalisation)** 測量程序與理論建構之間的邏輯關連條件，或是將理論建構轉換為測量程序的過程。
- **變項(variable)** 執行測量程序的最終成品就是“資料”，一個變項就是"資料”的集合[^02-transltion-03]。

[^02-transltion-02]: 譯注~根據名詞解說，此處中譯為”測量程序“。

[^02-transltion-03]: 譯注~本書的多數單元將variable翻成"變項"，除了統計理論的單元大多會翻成“變數”。因為"變項"是資料的，“變數”是數學的。

在統計實務，即使是訓練過的科學家也不大會去管這些名詞的差異，但是對於正在學習的同學們，搞清楚這些名詞的涵義會很有幫助。

<!---All of the ideas discussed in the previous section relate to the concept of **operationalisation**. To be a bit more precise about the idea, operationalisation is the process by which we take a meaningful but somewhat vague concept and turn it into a precise measurement. The process of operationalisation can involve several different things:

-   Being precise about what you are trying to measure. For instance, does "age" mean "time since birth" or "time since conception" in the context of your research?

-   Determining what method you will use to measure it. Will you use self-report to measure age, ask a parent, or look up an official record? If you're using self-report, how will you phrase the question?

-   Defining the set of allowable values that the measurement can take. Note that these values don't always have to be numerical, though they often are. When measuring age the values are numerical, but we still need to think carefully about what numbers are allowed. Do we want age in years, years and months, days, or hours? For other types of measurements (e.g., gender) the values aren't numerical. But, just as before, we need to think about what values are allowed. If we're asking people to self-report their gender, what options to we allow them to choose between? Is it enough to allow only "male" or "female"? Do you need an "other" option? Or should we not give people specific options and instead let them answer in their own words? And if you open up the set of possible values to include all verbal response, how will you interpret their answers?

Operationalisation is a tricky business, and there's no "one, true way" to do it. The way in which you choose to operationalise the informal concept of "age" or "gender" into a formal measurement depends on what you need to use the measurement for. Often you'll find that the community of scientists who work in your area have some fairly well-established ideas for how to go about it. In other words, operationalisation needs to be thought through on a case by case basis. Nevertheless, while there a lot of issues that are specific to each individual research project, there are some aspects to it that are pretty general.

Before moving on I want to take a moment to clear up our terminology, and in the process introduce one more term. Here are four different things that are closely related to each other:

-   **A theoretical construct.** This is the thing that you're trying to take a measurement of, like "age", "gender" or an "opinion". A theoretical construct can't be directly observed, and often they're actually a bit vague.
-   **A measure.** The measure refers to the method or the tool that you use to make your observations. A question in a survey, a behavioural observation or a brain scan could all count as a measure.
-   **An operationalisation.** The term "operationalisation" refers to the logical connection between the measure and the theoretical construct, or to the process by which we try to derive a measure from a theoretical construct.
-   **A variable.** Finally, a new term. A variable is what we end up with when we apply our measure to something in the world. That is, variables are the actual "data" that we end up with in our data sets.

In practice, even scientists tend to blur the distinction between these things, but it's very helpful to try to understand the differences.--->

## 測量尺度 {#sec-Scales-of-measurement}

[前一節](02-A-brief-introduction-to-research-design.html#sec-Introduction-to-psychological-measurement)最後提到心理測量程序的產物就是變項。一份資料檔通常不只有一個變項，而且各種變項有本質的區別，所以這一節要好好認識幾種變項型態。同學們將認識各種**測量尺度**以及學會判斷資料的變項型態。

<!---As the previous section indicates, the outcome of a psychological measurement is called a variable. But not all variables are of the same qualitative type and so it's useful to understand what types there are. A very useful concept for distinguishing between different types of variables is what's known as **scales of measurement**. --->

### 名義尺度

**名義尺度**變項又稱**類別**變項，命名理由是代表資料的每個數值之間沒有任何明確關係。名義尺度變項的資料之間，沒有那一個數值比較"大"或比較"好"，因此也無法計算這筆變項的平均值。最典型的例子是"瞳孔顏色"。人類瞳孔有藍色、綠色、棕色等各種顏色。沒有人能說瞳孔有所謂的"平均顏色"。人類性別也是典型的名義尺度變項：男性或女性沒有那一種比較好或比較不好，當然也沒有所謂的"平均性別"。總而言之，以名義尺度紀錄的資料，不同數值之間只有各自不一樣的意思而已。

讓我們再用一個例子更深入了解名義尺度。假如今天要研究民眾如何通勤上班，我們可以設定資料變項記錄每個人是用什麼交通方式去上班。"通勤方式"這個變項可能有好幾種數值：像是“火車”，“公車”，“汽車”，“自行車”。假設調查100位民眾，得到了這四種回答，完成的紀錄就如表2-1。

<!---A **nominal scale** variable (also referred to as a **categorical** variable) is one in which there is no particular relationship between the different possibilities. For these kinds of variables it doesn't make any sense to say that one of them is "bigger' or "better" than any other one, and it absolutely doesn't make any sense to average them. The classic example for this is "eye colour". Eyes can be blue, green or brown, amongst other possibilities, but none of them is any "bigger" than any other one. As a result, it would feel really weird to talk about an "average eye colour". Similarly, gender is nominal too: male isn't better or worse than female. Neither does it make sense to try to talk about an "average gender". In short, nominal scale variables are those for which the only thing you can say about the different possibilities is that they are different. That's it.

Let's take a slightly closer look at this. Suppose I was doing research on how people commute to and from work. One variable I would have to measure would be what kind of transportation people use to get to work. This "transport type" variable could have quite a few possible values, including: "train", "bus", "car", "bicycle". For now, let's suppose that these four are the only possibilities. Then imagine that I ask 100 people how they got to work today, with this result (見表2-1)。(@tbl-tab2-1).--->


```{r}
#| label: tbl-tab2-1
#| tbl-cap: How did 100 people get to work today
#huxtabs[[2]][[1]] 
```

<center>表2-1: 調查100位民眾本日通勤方式</center>

|通勤方式|人數|
|:---:|:---:|
|(1)火車|12|
|(2)公車|30|
|(3)汽車|48|
|(4)自行車|10|

那麼這裡可以找出平均的通勤方式嗎？很明顯答案不只一個，或者說這是個蠢問題。也許你會說汽車是最多人上班的交通工具，也可以說火車最沒有民眾使用，但都不能涵括調查結果。同學們也可以看一看表2-2，無論如何變動記錄表裡的項目順序，都無法讓這筆資料透露任何意義。

<!---So, what's the average transportation type? Obviously, the answer here is that there isn't one. It's a silly question to ask. You can say that travel by car is the most popular method, and travel by train is the least popular method, but that's about all. Similarly, notice that the order in which I list the options isn't very interesting. I could have chosen to display the data like in 表2-2。 @tbl-tab2-2.--->

```{r}
#| label: tbl-tab2-2
#| tbl-cap: How did 100 people get to work today, a different view
#huxtabs[[2]][[2]]
```


<center>表2-2: 調查100位民眾本日通勤方式，依上段描述更改呈現順序</center>

|通勤方式|人數|
|:---:|:---:|
|(3)汽車|48|
|(1)火車|12|
|(4)自行車|10|
|(2)公車|30|

<!---...and nothing really changes.--->

...很明顯沒有什麼改變。

### 次序尺度

**次序尺度**變項比起名義尺度變項能呈現一點有結構的資訊，不過沒有很多。次序尺度變項使用自然有意義的方式排序資料數值，但是除了排序我們就無法再做什麼了。最典型的例子是“賽跑選手抵達終點的名次”。先抵達終點的選手毫無疑問比第二名選手快，不過紀錄看不出快多少。紀錄一場比賽的選手名次後，我們會知道第一名比第二名快，第二名比第三名快，但是從資料看不出第一名和第二名的差距，與第二名和第三名的差距有什麼差異。

以下是個比較有心理學意義的例子。假如我想調查民眾對於氣候變遷的態度，我設計了以下幾個項目，請受訪者選出其中一個最接近個人看法的項目。

1. 氣溫確實有上升，而且是人類活動造成的
2. 氣溫確實有上升，但是原因不明
3. 氣溫確實有上升，但是與人類活動無關
4. 氣溫沒有變化

以上四個項目的順序是根據當前已知的科學研究資訊排列，第1項看法最符合現在的研究所知，第2與第3項是尚有研究證據支持的看法，第4項看法與所有研究證據相左。若是受測者們對於已知科學證據有充分了解，以上四個項目的排序就是符合多數人的看法。如果我將選項用以下段落的方式排序，則是不符合多數人的看法。

1. 氣溫確實有上升，但是與人類活動無關
2. 氣溫確實有上升，而且是人類活動造成的
3. 氣溫沒有變化
4. 氣溫確實有上升，但是原因不明

...這個例子說明心理測量的“結構”是資料呈現出多數人的合理回應。

假設我成功收集了100位民眾的回應，調查結果總計如表2-3。

<!---**Ordinal scale** variables have a bit more structure than nominal scale variables, but not by a lot. An ordinal scale variable is one in which there is a natural, meaningful way to order the different possibilities, but you can't do anything else. The usual example given of an ordinal variable is "finishing position in a race". You *can* say that the person who finished first was faster than the person who finished second, but you *don't* know how much faster. As a consequence we know that 1st $>$ 2nd, and we know that 2nd $>$ 3rd, but the difference between 1st and 2nd might be much larger than the difference between 2nd and 3rd.

Here's a more psychologically interesting example. Suppose I'm interested in people's attitudes to climate change. I then go and ask some people to pick the statement (from four listed statements) that most closely matches their beliefs:

1.  Temperatures are rising because of human activity
2.  Temperatures are rising but we don't know why
3.  Temperatures are rising but not because of humans
4.  Temperatures are not rising

Notice that these four statements actually do have a natural ordering, in terms of "the extent to which they agree with the current science". Statement 1 is a close match, statement 2 is a reasonable match, statement 3 isn't a very good match, and statement 4 is in strong opposition to current science. So, in terms of the thing I'm interested in (the extent to which people endorse the science), I can order the items as 1 $>$ 2 $>$ 3 $>$ 4. Since this ordering exists, it would be very weird to list the options like this...

1.  Temperatures are rising but not because of humans
2.  Temperatures are rising because of human activity
3.  Temperatures are not rising
4.  Temperatures are rising but we don't know why

...because it seems to violate the natural "structure" to the question.

So, let's suppose I asked 100 people these questions, and got the answers shown in 表2-3。 @tbl-tab2-3. --->

```{r}
#| label: tbl-tab2-3
#| tbl-cap: Attitudes to climate change
#huxtabs[[2]][[3]]
```

<center>表2-3: 氣候變遷態度調查結果</center>

|調查項目回應|人數|
|:---:|:---:|
|(1)氣溫確實有上升，而且是人類活動造成的|51|
|(2)氣溫確實有上升，但是原因不明|20|
|(3)氣溫確實有上升，但是與人類活動無關|10|
|(4)氣溫沒有變化|19|

資料分析能獲得幾種合理解釋，其中一種是最多民眾回應代表理性意見的(1),(2),(3)，或者說100位受訪中的81人*至少有理解*最新科學研究。另一種合理解釋是至少一半受訪者*傾向不同意*氣候變遷是現實的看法，因為100位受訪者中的49人選擇不同於主流科學觀點的(2),(3), (4)。然而也可能很難合理解釋為何有九成受訪者選擇(1),(2),(4)，因為這三項的排序不符合預期的順序結構，無法說明最多人選擇的三項有什麼意義。

以上的說明是向同學們示範，次序尺度變項資料要符合預期中的”順序結構“，才會有合理的分析，而且我們不能計算數值之間的平均。如果我們用回應人數計算加權平均，雖然可以得到平均值1.97，但是這個數值對於解釋調查結果沒有任何幫助。請同學想想這個數字能不能做為報告的結論。

<!---When analysing these data it seems quite reasonable to try to group (1), (2) and (3) together, and say that 81 out of 100 people were willing to at *least partially* endorse the science. And it's also quite reasonable to group (2), (3) and (4) together and say that 49 out of 100 people registered *at least some disagreement* with the dominant scientific view. However, it would be entirely bizarre to try to group (1), (2) and (4) together and say that 90 out of 100 people said... what? There's nothing sensible that allows you to group those responses together at all.

That said, notice that while we *can* use the natural ordering of these items to construct sensible groupings, what we can't do is average them. For instance, in my simple example here, the "average" response to the question is 1.97. If you can tell me what that means I'd love to know, because it seems like gibberish to me! --->



### 等距尺度

不同於名義尺度與次序尺度變項，**等距尺度**和比例尺度變項以可數的數字表數值，能獲得有意義的資訊。等距尺度變項的*數值之間差異*是可以推論的，但是變項數值沒有“自然的”零點。攝氏溫度是一個說明等距尺度變項的好例子。例如是，昨天氣溫是 15$^{\circ}$，今天氣溫是 18$^{\circ}$，所以兩日的溫差是3$^{\circ}$。更重要的是，這個3$^{\circ}$與氣溫7$^{\circ}$ 與 10$^{\circ}$的差異是*完全相等*。簡言之，用等距尺度變項數值的做加法或減法是有意義的[^02-a-brief-introduction-to-research-design-2]。

[^02-a-brief-introduction-to-research-design-2]:<原作者的補充>其實曾有物理學大神告訴我，溫度嚴格來說不是等距尺度變項，因為只要環境溫度至少有3$^{\circ}$ ，任何物體都可以加熱。雖然在物理學來說，這並不是一個講解等距尺度好例子，不過為了方便說明，這裡暫且無視物理事實。

然而，攝氏零度並不是“量不到溫度”的意思，最早攝氏零度是根據“觀察到水開始結冰”而指定的數值。這造成溫度的數值無法相乘或相除：要說20$^{\circ}$ 比10$^{\circ}$ 熱兩倍是荒誔的，也不可能說20$^{\circ}$ 是-10$^{\circ}$ 的負兩倍。

我們再來舉心理學的例子吧。假如我想要了解各位同學在大學四年間學習態度的變化，最好的方式是每學年或學期開始，就請同手們做一次態度調查。這樣的記錄就是等距尺度變項資料。如果我手上的紀錄有一位學生是2013年的，另一位學生的紀錄是2018年的，向你們報告說2018年學生接受調查的時間是2013年的"1.0025倍“，這樣說奇不奇怪呢？


<!---In contrast to nominal and ordinal scale variables, **interval scale** and ratio scale variables are variables for which the numerical value is genuinely meaningful. In the case of interval scale variables the *differences* between the numbers are interpretable, but the variable doesn't have a "natural" zero value. A good example of an interval scale variable is measuring temperature in degrees celsius. For instance, if it was 15$^{\circ}$ yesterday and 18$^{\circ}$ today, then the 3$^{\circ}$ difference between the two is genuinely meaningful. Moreover, that 3$^{\circ}$ difference is *exactly the same* as the 3$^{\circ}$ difference between 7$^{\circ}$ and 10$^{\circ}$. In short, addition and subtraction are meaningful for interval scale variables.[^02-a-brief-introduction-to-research-design-2]

[^02-a-brief-introduction-to-research-design-2]: Actually, I've been informed by readers with greater physics knowledge than I that temperature isn't strictly an interval scale, in the sense that the amount of energy required to heat something up by 3° depends on it's current temperature. So in the sense that physicists care about, temperature isn't actually an interval scale. But it still makes a cute example so I'm going to ignore this little inconvenient truth.

However, notice that the 0$^{\circ}$ does not mean "no temperature at all". It actually means "the temperature at which water freezes", which is pretty arbitrary. As a consequence it becomes pointless to try to multiply and divide temperatures. It is wrong to say that 20$^{\circ}$ is twice as hot as 10$^{\circ}$, just as it is weird and meaningless to try to claim that 20$^{\circ}$ is negative two times as hot as -10$^{\circ}$.

Again, lets look at a more psychological example. Suppose I'm interested in looking at how the attitudes of first-year university students have changed over time. Obviously, I'm going to want to record the year in which each student started. This is an interval scale variable. A student who started in 2003 did arrive 5 years before a student who started in 2008. However, it would be completely daft for me to divide 2008 by 2003 and say that the second student started "1.0024 times later" than the first one. That doesn't make any sense at all. --->

### 比例尺度

最後一種變項型態是有零點的**比例尺度**，也就是說變項數值是可以相乘且相除的。有個不錯的心理學研究常用的比例尺度變項是反應時間(RT)。許多心理學作業都會紀錄參與者花了多少時間解決問題或給出回答，因為時間長度是作業難度的指標。假如今天有項作業，參與者A花了2.3秒回答，另一位參與者B花了3.1秒。就像等距尺度變項，比例尺度數值的相加與相減是有意義的，也就是我們可以說受試者B比起受試者A多花了3.1 - 2.3 = 0.8秒。同學也要留意反應時間數值的乘法與除法是有意義的：我們可以說受試者B比受試者A花了 3.1/2.3 = 1.35倍的時間完成回答。能做完整四則運算的原因是，反應時間有真正的"零點”~零秒就是沒有反應記錄。

<!---The fourth and final type of variable to consider is a **ratio scale** variable, in which zero really means zero, and it's okay to multiply and divide. A good psychological example of a ratio scale variable is response time (RT). In a lot of tasks it's very common to record the amount of time somebody takes to solve a problem or answer a question, because it's an indicator of how difficult the task is. Suppose that Alan takes 2.3 seconds to respond to a question, whereas Ben takes 3.1 seconds. As with an interval scale variable, addition and subtraction are both meaningful here. Ben really did take 3.1 - 2.3 = 0.8 seconds longer than Alan did. However, notice that multiplication and division also make sense here too: Ben took 3.1/2.3 = 1.35 times as long as Alan did to answer the question. And the reason why you can do this is that for a ratio scale variable such as RT "zero seconds" really does mean "no time at all". --->

### 連續與間斷變項

還有另一套變項型態是同學們需要知道的：連續變項與間斷變項。不論你要處理的測量尺度是什麼，也會具備**切換**為連續變項與間斷變項的條件 (見**表2-4**)。我將連續與間斷變項的差異整理一下：

- **連續變項(continuous variable)**的任何兩個數值之間，都能存在另一個數值，因此是連續的。
- **間斷變項(discrete variable)**的數值當然不是連續的。任何兩個相鄰的間斷變項數值，不可能存在其他數值。

<!---There's a second kind of distinction that you need to be aware of, regarding what types of variables you can run into. This is the distinction between continuous variables and discrete variables (見表2-4)。(@tbl-tab2-4). The difference between these is as follows:

-   A **continuous variable** is one in which, for any two values that you can think of, it's always logically possible to have another value in between.
-   A **discrete variable** is, in effect, a variable that isn't continuous. For a discrete variable it's sometimes the case that there's nothing in the middle. --->

```{r}
#| label: tbl-tab2-4
#| tbl-cap: The relationship between the scales of measurement and the discrete/continuity distinction. Cells with a tick mark correspond to things that are possible
#huxtabs[[2]][[4]] 
```

<center>表2-4: 測量尺度與間斷/連續變項的關聯性。細格內的符號是jamovi的變項標示符號。[^02-translation-2]</center>

||連續變項|間斷變項|
|:---:|:---:|:---:|
|名義尺度||![](images/variable-nominal.svg),![](images/variable-id.svg)|
|次序尺度||![](images/variable-ordinal.svg)|
|等距尺度|![](images/variable-continuous.svg)|![](images/variable-nominal.svg),![](images/variable-ordinal.svg)|
|比例尺度|![](images/variable-continuous.svg)|![](images/variable-nominal.svg),![](images/variable-ordinal.svg)|

[^02-translation-2]: 譯註~原著並未提供變項符號，表內符號取自[jamovi說明文件](https://docs.jamovi.org/_pages/um_4_spreadsheet.html)。

雖然表2-5看起來有點抽象，透過一些例子就能理解如何切換。同樣用解釋比例尺度的反應時間為例，現在除了有參與者A用了2.3秒，與參與者B用了3.1秒的資料，還有一位參與者C用了3.0秒鐘，剛好記錄在前兩位之間。當然，若再有一位參與者D的記錄是3.031秒，就是在B與C之間有一筆資料。雖然真正的實驗不一定會測量得如此準確，這只是示範連續變項的主要特性是在已經存在的任何兩筆資料數值之間，都能增加一個新的資料數值。

只要變項無法在任何兩個資料數值之間增加資料，就只能是間斷變項。像是名義尺度變項永遠都只能是間斷變項。如同火車走的鐵路與自行車道之間不可能有“切換機制”，名義數值2與3之間不可能增加一個數值2.3，所以名義尺度變項資料只能當成間斷變項資料處理。類別尺度變項也是只能切換成間斷變項，雖然"第2名"確實是在”第1名"與"第3名"之間，"第1名"與"第2名"之間沒有空間給其他數值。至於等距尺度與比例尺度變項，可以切換為間斷變項，也可以切換為連續變項。前面提到的反應時間(比例尺度)與攝氏溫度(等距尺度)都是可以切換為連續變項。不過如果是各位同學的入學年份，雖然是等距尺度變項，卻只能切換為間斷變項，像是2022年與2023年之間不能放入其他年份。還有假如今天同學們做了一份都是是非題的測驗，雖然每一題分數是等距尺度變項資料，也是只能切換為間斷變項，因為沒有5/10正確或6/10錯誤之類的回答。**表2-4**總結四種測量尺度變項在jamovi介面標記符號，以及能否切換為間斷變項或連續變項[^02-translation-3]。這一節特別解釋各種測量尺度與間斷/連續的切換關係，出於原作者的兩個理由：首先是有些統計教科書混淆了測量尺度與間斷/連續變項的定義，其次是經常聽到很多人，包括資深研究人員與統計學教師，提到「間斷變項」都會直覺認定是「名義尺度變項」，理解清楚的話就會知道這樣認定會誤讀資料[^02-translation-4]。

[^02-translation-3]: 譯註~原文的**表2-4**並未說明jamovi符號，改編是為了幫助讀者了解後續章節的操作說明，建議下載[示範檔案](files/Variable_types.omv)並搭配[中文版示範影片](https://youtu.be/_8X9m-aU-tA)閱讀。

[^02-translation-4]: 譯註~中文化作者認為還有第三個理由。由於有些統計教程混淆測量尺度與間斷/連續變項的切換關係，造成學習者低估機率論在學習統計的重要性。在[第7章](07-Introduction-to-probability.html)，我們會了解間斷/連續變項的數學意義。


<!---These definitions probably seem a bit abstract, but they're pretty simple once you see some examples. For instance, response time is continuous. If Alan takes 3.1 seconds and Ben takes 2.3 seconds to respond to a question, then Cameron's response time will lie in between if he took 3.0 seconds. And of course it would also be possible for David to take 3.031 seconds to respond, meaning that his RT would lie in between Cameron's and Alan's. And while in practice it might be impossible to measure RT that precisely, it's certainly possible in principle. Because we can always find a new value for RT in between any two other ones we regard RT as a continuous measure.

Discrete variables occur when this rule is violated. For example, nominal scale variables are always discrete. There isn't a type of transportation that falls "in between" trains and bicycles, not in the strict mathematical way that 2.3 falls in between 2 and 3. So transportation type is discrete. Similarly, ordinal scale variables are always discrete. Although "2nd place" does fall between "1st place" and "3rd place", there's nothing that can logically fall in between "1st place" and "2nd place". Interval scale and ratio scale variables can go either way. As we saw above, response time (a ratio scale variable) is continuous. Temperature in degrees celsius (an interval scale variable) is also continuous. However, the year you went to school (an interval scale variable) is discrete. There's no year in between 2002 and 2003. The number of questions you get right on a true-or-false test (a ratio scale variable) is also discrete. Since a true-or-false question doesn't allow you to be "partially correct", there's nothing in between 5/10 and 6/10. @tbl-tab2-4 summarises the relationship between the scales of measurement and the discrete/continuity distinction. Cells with a tick mark correspond to things that are possible. I'm trying to hammer this point home, because (a) some textbooks get this wrong, and (b) people very often say things like "discrete variable" when they mean "nominal scale variable". It's very unfortunate. --->

### 複雜的現實

好啦，測量尺度與資料變項之間的切換規律或許會讓一些讀者感到震驚，不過真實世界遠比這條小規律複雜許多。其實現實生活中，只有非常少的可測量指標充分符合這條規律，所以同學們必須留意，不要把切換規律硬套在任何你遇到的測量尺度。切換規律只是指引而已，提示統計使用者在實務中如何找出處置資料的最佳方式而已[^02-translation-5]。

[^02-translation-5]: 譯註~到本書最新版上線的時間，還沒有一種有效的應用程式或人工智慧程序能幫助人類判斷適用分析任務的資料變項型態，請參考這份[示範檔案](files/Variable_types.omv)的[中文版示範影片](https://youtu.be/_8X9m-aU-tA)。

讓我們用一個心理學家滿常用的心理測量工具～**李克特量表(Likert scale)**來說明現實世界有多複雜。許多調查問卷都會使用李克特量表收集受測者的回應，讀者與同學們也許已經填過好多份使用李克特量表的問卷，搞不好曾經在某份自行設計的問卷使用過李克特量表。以下是一條假想的問卷問題：

請從以下五個選項，挑出最符合您對於*「所有韓星都很潮」*[^02-translation-6]這句描述的看法？

[^02-translation-6]: 譯註~原文的題目不是這一句，中文版改成接地氣的範例。

1. 完全不同意
2. 部分不同意
3. 即非同意也非不同意
4. 部分同意
5. 完全同意

這是典型的五點式李克特量表，依數值大小順序排列同意程度，讓受測者選一個數字，通常每個數字旁邊都有文字說明。不過也不必所有數字旁邊都要放文字說明，所以問卷選項也能寫成這個樣子：

1. 完全不同意
2. 
3. 
4. 
5. 完全同意

李克特量表是非常好用，但是用途有限的工具。怎麼說？請同學想想用這個問題收集到的回應是什麼樣的資料變項？很明顯應該是間斷變項，因為沒有人能給2.5這種答案。這筆資料也顯然不是名義尺度，因為選項是有順序的；也顯然不是比例尺度，因為沒有自然零點。

那麼這筆資料是次序尺度還是等距尺度？有一種說法是我們無法確定"部分同意"和"完全同意"之間的數值差異，與"部分同意"和"即非同意也非不同意"之間的數值差異是相等的。其實拿日常生活的任何事物做成調查問題，不必有數學知識，任何人都能同意李克特量表的任何一對相鄰數字之間差異是不相等的。所以說我們不應該將用李克特量表收集的資料當成次序尺度變項。另一種說法是假設受測者填答時，會以為選項數字1到5是均等平分的量尺，心理預設五個選項之間的差異如同標示在選項前的數字一樣。經常使用到今天，多數研究者都將李克特量表測得的資料當成等距尺度[^02-a-brief-introduction-to-research-design-3]。但是嚴格來說又不能算是等距尺度，所以在實務上心理學研究者常將李克特量表當成**準等距尺度**。

[^02-a-brief-introduction-to-research-design-3]: 唉，心理學的問題從來沒有一種能簡單回答的方式！

<!---Okay, I know you're going to be shocked to hear this, but the real world is much messier than this little classification scheme suggests. Very few variables in real life actually fall into these nice neat categories, so you need to be kind of careful not to treat the scales of measurement as if they were hard and fast rules. It doesn't work like that. They're guidelines, intended to help you think about the situations in which you should treat different variables differently. Nothing more.

So let's take a classic example, maybe *the* classic example, of a psychological measurement tool: the **Likert scale**. The humble Likert scale is the bread and butter tool of all survey design. You yourself have filled out hundreds, maybe thousands, of them and odds are you've even used one yourself. Suppose we have a survey question that looks like this:

Which of the following best describes your opinion of the statement that "all pirates are freaking awesome"?

and then the options presented to the participant are these:

1.  Strongly disagree
2.  Disagree
3.  Neither agree nor disagree
4.  Agree
5.  Strongly agree

This set of items is an example of a 5-point Likert scale, in which people are asked to choose among one of several (in this case 5) clearly ordered possibilities, generally with a verbal descriptor given in each case. However, it's not necessary that all items are explicitly described. This is a perfectly good example of a 5-point Likert scale too:

1.  Strongly disagree
2.  
3.  
4.  
5.  Strongly agree

Likert scales are very handy, if somewhat limited, tools. The question is what kind of variable are they? They're obviously discrete, since you can't give a response of 2.5. They're obviously not nominal scale, since the items are ordered; and they're not ratio scale either, since there's no natural zero.

But are they ordinal scale or interval scale? One argument says that we can't really prove that the difference between "strongly agree" and "agree" is of the same size as the difference between "agree" and "neither agree nor disagree". In fact, in everyday life it's pretty obvious that they're not the same at all. So this suggests that we ought to treat Likert scales as ordinal variables. On the other hand, in practice most participants do seem to take the whole "on a scale from 1 to 5" part fairly seriously, and they tend to act as if the differences between the five response options were fairly similar to one another. As a consequence, a lot of researchers treat Likert scale data as interval scale.[^02-a-brief-introduction-to-research-design-3] It's not interval scale, but in practice it's close enough that we usually think of it as being **quasi-interval scale**.

[^02-a-brief-introduction-to-research-design-3]: Ah, psychology... never an easy answer to anything! --->

## 測量的信度 {#sec-Assessing-the-reliability-of-a-measurement}

At this point we've thought a little bit about how to operationalise a theoretical construct and thereby create a psychological measure. And we've seen that by applying psychological measures we end up with variables, which can come in many different types. At this point, we should start discussing the obvious question: is the measurement any good? We'll do this in terms of two related ideas: *reliability* and *validity*. Put simply, the **reliability** of a measure tells you how precisely you are measuring something, whereas the validity of a measure tells you how accurate the measure is. In this section I'll talk about reliability; we'll talk about validity in in the section on [Assessing the validity of a study].

Reliability is actually a very simple concept. It refers to the repeatability or consistency of your measurement. The measurement of my weight by means of a "bathroom scale" is very reliable. If I step on and off the scales over and over again, it'll keep giving me the same answer. Measuring my intelligence by means of "asking my mum" is very unreliable. Some days she tells me I'm a bit thick, and other days she tells me I'm a complete idiot. Notice that this concept of reliability is different to the question of whether the measurements are correct (the correctness of a measurement relates to it's validity). If I'm holding a sack of potatos when I step on and off the bathroom scales the measurement will still be reliable: it will always give me the same answer. However, this highly reliable answer doesn't match up to my true weight at all, therefore it's wrong. In technical terms, this is a reliable but invalid measurement. Similarly, whilst my mum's estimate of my intelligence is a bit unreliable, she might be right. Maybe I'm just not too bright, and so while her estimate of my intelligence fluctuates pretty wildly from day to day, it's basically right. That would be an unreliable but valid measure. Of course, if my mum's estimates are too unreliable it's going to be very hard to figure out which one of her many claims about my intelligence is actually the right one. To some extent, then, a very unreliable measure tends to end up being invalid for practical purposes; so much so that many people would say that reliability is necessary (but not sufficient) to ensure validity.

Okay, now that we're clear on the distinction between reliability and validity, let's have a think about the different ways in which we might measure reliability:

-   **Test-retest reliability**. This relates to consistency over time. If we repeat the measurement at a later date do we get a the same answer?
-   **Inter-rater reliability**. This relates to consistency across people. If someone else repeats the measurement (e.g., someone else rates my intelligence) will they produce the same answer?
-   **Parallel forms reliability**. This relates to consistency across theoretically-equivalent measurements. If I use a different set of bathroom scales to measure my weight does it give the same answer?
-   **Internal consistency reliability**. If a measurement is constructed from lots of different parts that perform similar functions (e.g., a personality questionnaire result is added up across several questions) do the individual parts tend to give similar answers. We'll look at this particular form of reliability later in the book, in @sec-Internal-consistency-reliability-analysis.

Not all measurements need to possess all forms of reliability. For instance, educational assessment can be thought of as a form of measurement. One of the subjects that I teach, *Computational Cognitive Science*, has an assessment structure that has a research component and an exam component (plus other things). The exam component is *intended* to measure something different from the research component, so the assessment as a whole has low internal consistency. However, within the exam there are several questions that are intended to (approximately) measure the same things, and those tend to produce similar outcomes. So the exam on its own has a fairly high internal consistency. Which is as it should be. You should only demand reliability in those situations where you want to be measuring the same thing!

## 變項的"角色”: 預測變項與結果變項

I've got one last piece of terminology that I need to explain to you before moving away from variables. Normally, when we do some research we end up with lots of different variables. Then, when we analyse our data, we usually try to explain some of the variables in terms of some of the other variables. It's important to keep the two roles "thing doing the explaining" and "thing being explained" distinct. So let's be clear about this now. First, we might as well get used to the idea of using mathematical symbols to describe variables, since it's going to happen over and over again. Let's denote the "to be explained" variable $Y$, and denote the variables "doing the explaining" as $X_1 , X_2$, etc.

When we are doing an analysis we have different names for $X$ and $Y$, since they play different roles in the analysis. The classical names for these roles are **independent variable** (IV) and **dependent variable** (DV). The IV is the variable that you use to do the explaining (i.e., $X$) and the DV is the variable being explained (i.e.,\$Y \$). The logic behind these names goes like this: if there really is a relationship between $X$ and $Y$ then we can say that $Y$depends on $X$, and if we have designed our study "properly" then $X$ isn't dependent on anything else. However, I personally find those names horrible. They're hard to remember and they're highly misleading because (a) the IV is never actually "independent of everything else", and (b) if there's no relationship then the DV doesn't actually depend on the IV. And in fact, because I'm not the only person who thinks that IV and DV are just awful names, there are a number of alternatives that I find more appealing. The terms that I'll use in this book are **predictors** and **outcomes**. The idea here is that what you're trying to do is use $X$ (the predictors) to make guesses about $Y$ (the outcomes).[^02-a-brief-introduction-to-research-design-4] This is summarised in @tbl-tab2-5.

[^02-a-brief-introduction-to-research-design-4]: Annoyingly though, there's a lot of different names used out there. I won't list all of them -- there would be no point in doing that -- other than to note that "response variable" is sometimes used where I've used "outcome". Sigh. This sort of terminological confusion is very common, I'm afraid.

```{r}
#| label: tbl-tab2-5
#| tbl-cap: Variable distinctions
huxtabs[[2]][[5]] 
```

## 實驗與非實驗研究

One of the big distinctions that you should be aware of is the distinction between "experimental research" and "non-experimental research". When we make this distinction, what we're really talking about is the degree of control that the researcher exercises over the people and events in the study.

### 實驗研究

The key feature of **experimental research** is that the researcher controls all aspects of the study, especially what participants experience during the study. In particular, the researcher manipulates or varies the predictor variables (IVs) but allows the outcome variable (DV) to vary naturally. The idea here is to deliberately vary the predictors (IVs) to see if they have any causal effects on the outcomes. Moreover, in order to ensure that there's no possibility that something other than the predictor variables is causing the outcomes, everything else is kept constant or is in some other way "balanced", to ensure that they have no effect on the results. In practice, it's almost impossible to *think* of everything else that might have an influence on the outcome of an experiment, much less keep it constant. The standard solution to this is **randomisation**. That is, we randomly assign people to different groups, and then give each group a different treatment (i.e., assign them different values of the predictor variables). We'll talk more about randomisation later, but for now it's enough to say that what randomisation does is minimise (but not eliminate) the possibility that there are any systematic difference between groups.

Let's consider a very simple, completely unrealistic and grossly unethical example. Suppose you wanted to find out if smoking causes lung cancer. One way to do this would be to find people who smoke and people who don't smoke and look to see if smokers have a higher rate of lung cancer. This is *not* a proper experiment, since the researcher doesn't have a lot of control over who is and isn't a smoker. And this really matters. For instance, it might be that people who choose to smoke cigarettes also tend to have poor diets, or maybe they tend to work in asbestos mines, or whatever. The point here is that the groups (smokers and non-smokers) actually differ on lots of things, not just smoking. So it might be that the higher incidence of lung cancer among smokers is caused by something else, and not by smoking per se. In technical terms these other things (e.g. diet) are called "confounders", and we'll talk about those in just a moment.

In the meantime, let's consider what a proper experiment might look like. Recall that our concern was that smokers and non-smokers might differ in lots of ways. The solution, as long as you have no ethics, is to control who smokes and who doesn't. Specifically, if we randomly divide young non-smokers into two groups and force half of them to become smokers, then it's very unlikely that the groups will differ in any respect other than the fact that half of them smoke. That way, if our smoking group gets cancer at a higher rate than the non-smoking group, we can feel pretty confident that (a) smoking does cause cancer and (b) we're murderers.

### 非實驗研究

**Non-experimental research** is a broad term that covers "any study in which the researcher doesn't have as much control as they do in an experiment". Obviously, control is something that scientists like to have, but as the previous example illustrates there are lots of situations in which you can't or shouldn't try to obtain that control. Since it's grossly unethical (and almost certainly criminal) to force people to smoke in order to find out if they get cancer, this is a good example of a situation in which you really shouldn't try to obtain experimental control. But there are other reasons too. Even leaving aside the ethical issues, our "smoking experiment" does have a few other issues. For instance, when I suggested that we "force" half of the people to become smokers, I was talking about *starting* with a sample of non-smokers, and then forcing them to become smokers. While this sounds like the kind of solid, evil experimental design that a mad scientist would love, it might not be a very sound way of investigating the effect in the real world. For instance, suppose that smoking only causes lung cancer when people have poor diets, and suppose also that people who normally smoke do tend to have poor diets. However, since the "smokers" in our experiment aren't "natural" smokers (i.e., we forced non-smokers to become smokers, but they didn't take on all of the other normal, real life characteristics that smokers might tend to possess) they probably have better diets. As such, in this silly example they wouldn't get lung cancer and our experiment will fail, because it violates the structure of the "natural" world (the technical name for this is an "artefactual" result).

One distinction worth making between two types of non-experimental research is the difference between **quasi-experimental research** and **case studies**. The example I discussed earlier, in which we wanted to examine incidence of lung cancer among smokers and non-smokers without trying to control who smokes and who doesn't, is a quasi-experimental design. That is, it's the same as an experiment but we don't control the predictors (IVs). We can still use statistics to analyse the results, but we have to be a lot more careful and circumspect.

The alternative approach, case studies, aims to provide a very detailed description of one or a few instances. In general, you can't use statistics to analyse the results of case studies and it's usually very hard to draw any general conclusions about "people in general" from a few isolated examples. However, case studies are very useful in some situations. Firstly, there are situations where you don't have any alternative. Neuropsychology has this issue a lot. Sometimes, you just can't find a lot of people with brain damage in a specific brain area, so the only thing you can do is describe those cases that you do have in as much detail and with as much care as you can. However, there's also some genuine advantages to case studies. Because you don't have as many people to study you have the ability to invest lots of time and effort trying to understand the specific factors at play in each case. This is a very valuable thing to do. As a consequence, case studies can complement the more statistically-oriented approaches that you see in experimental and quasi-experimental designs. We won't talk much about case studies in this book, but they are nevertheless very valuable tools!

## 研究的效度

More than any other thing, a scientist wants their research to be "valid". The conceptual idea behind **validity** is very simple. Can you trust the results of your study? If not, the study is invalid. However, whilst it's easy to state, in practice it's much harder to check validity than it is to check reliability. And in all honesty, there's no precise, clearly agreed upon notion of what validity actually is. In fact, there are lots of different kinds of validity, each of which raises it's own issues. And not all forms of validity are relevant to all studies. I'm going to talk about five different types of validity:

-   Internal validity
-   External validity
-   Construct validity
-   Face validity
-   Ecological validity

First, a quick guide as to what matters here. (1) Internal and external validity are the most important, since they tie directly to the fundamental question of whether your study really works. (2) Construct validity asks whether you're measuring what you think you are. (3) Face validity isn't terribly important except insofar as you care about "appearances". (4) Ecological validity is a special case of face validity that corresponds to a kind of appearance that you might care about a lot.

### 內在效度

**Internal validity** refers to the extent to which you are able draw the correct conclusions about the causal relationships between variables. It's called "internal" because it refers to the relationships between things "inside" the study. Let's illustrate the concept with a simple example. Suppose you're interested in finding out whether a university education makes you write better. To do so, you get a group of first year students, ask them to write a 1000 word essay, and count the number of spelling and grammatical errors they make. Then you find some third-year students, who obviously have had more of a university education than the first-years, and repeat the exercise. And let's suppose it turns out that the third-year students produce fewer errors. And so you conclude that a university education improves writing skills. Right? Except that the big problem with this experiment is that the third-year students are older and they've had more experience with writing things. So it's hard to know for sure what the causal relationship is. Do older people write better? Or people who have had more writing experience? Or people who have had more education? Which of the above is the true cause of the superior performance of the third-years? Age? Experience? Education? You can't tell. This is an example of a failure of internal validity, because your study doesn't properly tease apart the causal relationships between the different variables.

### 外在效度

**External validity** relates to the **generalisability** or **applicability** of your findings. That is, to what extent do you expect to see the same pattern of results in "real life" as you saw in your study. To put it a bit more precisely, any study that you do in psychology will involve a fairly specific set of questions or tasks, will occur in a specific environment, and will involve participants that are drawn from a particular subgroup (disappointingly often it is college students!). So, if it turns out that the results don't actually generalise or apply to people and situations beyond the ones that you studied, then what you've got is a lack of external validity.

The classic example of this issue is the fact that a very large proportion of studies in psychology will use undergraduate psychology students as the participants. Obviously, however, the researchers don't care *only* about psychology students. They care about people in general. Given that, a study that uses only psychology students as participants always carries a risk of lacking external validity. That is, if there's something "special" about psychology students that makes them different to the general population in some relevant respect, then we may start worrying about a lack of external validity.

That said, it is absolutely critical to realise that a study that uses only psychology students does not necessarily have a problem with external validity. I'll talk about this again later, but it's such a common mistake that I'm going to mention it here. The external validity of a study is threatened by the choice of population if (a) the population from which you sample your participants is very narrow (e.g., psychology students), and (b) the narrow population that you sampled from is systematically different from the general population in some respect that is relevant to the *psychological phenomenon that you intend to study*. The italicised part is the bit that lots of people forget. It is true that psychology undergraduates differ from the general population in lots of ways, and so a study that uses only psychology students may have problems with external validity. However, if those differences aren't very relevant to the phenomenon that you're studying, then there's nothing to worry about. To make this a bit more concrete here are two extreme examples:

-   You want to measure "attitudes of the general public towards psychotherapy", but all of your participants are psychology students. This study would almost certainly have a problem with external validity.
-   You want to measure the effectiveness of a visual illusion, and your participants are all psychology students. This study is unlikely to have a problem with external validity

Having just spent the last couple of paragraphs focusing on the choice of participants, since that's a big issue that everyone tends to worry most about, it's worth remembering that external validity is a broader concept. The following are also examples of things that might pose a threat to external validity, depending on what kind of study you're doing:

-   People might answer a "psychology questionnaire" in a manner that doesn't reflect what they would do in real life.
-   Your lab experiment on (say) "human learning" has a different structure to the learning problems people face in real life.

### 建構效度

**Construct validity** is basically a question of whether you're measuring what you want to be measuring. A measurement has good construct validity if it is actually measuring the correct theoretical construct, and bad construct validity if it doesn't. To give a very simple (if ridiculous) example, suppose I'm trying to investigate the rates with which university students cheat on their exams. And the way I attempt to measure it is by asking the cheating students to stand up in the lecture theatre so that I can count them. When I do this with a class of 300 students 0 people claim to be cheaters. So I therefore conclude that the proportion of cheaters in my class is 0%. Clearly this is a bit ridiculous. But the point here is not that this is a very deep methodological example, but rather to explain what construct validity is. The problem with my measure is that while I'm trying to measure "the proportion of people who cheat" what I'm actually measuring is "the proportion of people stupid enough to own up to cheating, or bloody minded enough to pretend that they do". Obviously, these aren't the same thing! So my study has gone wrong, because my measurement has very poor construct validity.

### 表面效度

**Face validity** simply refers to whether or not a measure "looks like" it's doing what it's supposed to, nothing more. If I design a test of intelligence, and people look at it and they say "no, that test doesn't measure intelligence", then the measure lacks face validity. It's as simple as that. Obviously, face validity isn't very important from a pure scientific perspective. After all, what we care about is whether or not the measure *actually* does what it's supposed to do, not whether it *looks like* it does what it's supposed to do. As a consequence, we generally don't care very much about face validity. That said, the concept of face validity serves three useful pragmatic purposes:

-   Sometimes, an experienced scientist will have a "hunch" that a particular measure won't work. While these sorts of hunches have no strict evidentiary value, it's often worth paying attention to them. Because often times people have knowledge that they can't quite verbalise, so there might be something to worry about even if you can't quite say why. In other words, when someone you trust criticises the face validity of your study, it's worth taking the time to think more carefully about your design to see if you can think of reasons why it might go awry. Mind you, if you don't find any reason for concern, then you should probably not worry. After all, face validity really doesn't matter very much.
-   Often (very often), completely uninformed people will also have a "hunch" that your research is crap. And they'll criticise it on the internet or something. On close inspection you may notice that these criticisms are actually focused entirely on how the study "looks", but not on anything deeper. The concept of face validity is useful for gently explaining to people that they need to substantiate their arguments further.
-   Expanding on the last point, if the beliefs of untrained people are critical (e.g., this is often the case for applied research where you actually want to convince policy makers of something or other) then you have to care about face validity. Simply because, whether you like it or not, a lot of people will use face validity as a proxy for real validity. If you want the government to change a law on scientific psychological grounds, then it won't matter how good your studies "really" are. If they lack face validity you'll find that politicians ignore you. Of course, it's somewhat unfair that policy often depends more on appearance than fact, but that's how things go.

### 生態效度

**Ecological validity** is a different notion of validity, which is similar to external validity, but less important. The idea is that, in order to be ecologically valid, the entire set up of the study should closely approximate the real world scenario that is being investigated. In a sense, ecological validity is a kind of face validity. It relates mostly to whether the study "looks" right, but with a bit more rigour to it. To be ecologically valid the study has to look right in a fairly specific way. The idea behind it is the intuition that a study that is ecologically valid is more likely to be externally valid. It's no guarantee, of course. But the nice thing about ecological validity is that it's much easier to check whether a study is ecologically valid than it is to check whether a study is externally valid. A simple example would be eyewitness identification studies. Most of these studies tend to be done in a university setting, often with a fairly simple array of faces to look at, rather than a line up. The length of time between seeing the "criminal" and being asked to identify the suspect in the "line up" is usually shorter. The "crime" isn't real so there's no chance of the witness being scared, and there are no police officers present so there's not as much chance of feeling pressured. These things all mean that the study definitely lacks ecological validity. They might (but might not) mean that it also lacks external validity.

## 混淆、人為結果、以及各種降低效度的因素

If we look at the issue of validity in the most general fashion the two biggest worries that we have are *confounders* and artefacts. These two terms are defined in the following way:

-   **Confounder:** A confounder is an additional, often unmeasured variable[^02-a-brief-introduction-to-research-design-5] that turns out to be related to both the predictors and the outcome. The existence of confounders threatens the internal validity of the study because you can't tell whether the predictor causes the outcome, or if the confounding variable causes it.
-   **Artefact:** A result is said to be "artefactual" if it only holds in the special situation that you happened to test in your study. The possibility that your result is an artefact describes a threat to your external validity, because it raises the possibility that you can't generalise or apply your results to the actual population that you care about.

[^02-a-brief-introduction-to-research-design-5]: The reason why I say that it's unmeasured is that if you have measured it, then you can use some fancy statistical tricks to deal with the confounder. Because of the existence of these statistical solutions to the problem of confounders, we often refer to a confounder that we have measured and dealt with as a covariate. Dealing with covariates is a more advanced topic, but I thought I'd mention it in passing since it's kind of comforting to at least know that this stuff exists.

As a general rule confounders are a bigger concern for non-experimental studies, precisely because they're not proper experiments. By definition, you're leaving lots of things uncontrolled, so there's a lot of scope for confounders being present in your study. Experimental research tends to be much less vulnerable to confounders. The more control you have over what happens during the study, the more you can prevent confounders from affecting the results. With random allocation, for example, confounders are distributed randomly, and evenly, between different groups.

However, there are always swings and roundabouts and when we start thinking about artefacts rather than confounders the shoe is very firmly on the other foot. For the most part, artefactual results tend to be a concern for experimental studies than for non-experimental studies. To see this, it helps to realise that the reason that a lot of studies are non-experimental is precisely because what the researcher is trying to do is examine human behaviour in a more naturalistic context. By working in a more real-world context you lose experimental control (making yourself vulnerable to confounders), but because you tend to be studying human psychology "in the wild" you reduce the chances of getting an artefactual result. Or, to put it another way, when you take psychology out of the wild and bring it into the lab (which we usually have to do to gain our experimental control), you always run the risk of accidentally studying something different to what you wanted to study.

Be warned though. The above is a rough guide only. It's absolutely possible to have confounders in an experiment, and to get artefactual results with non-experimental studies. This can happen for all sorts of reasons, not least of which is experimenter or researcher error. In practice, it's really hard to think everything through ahead of time and even very good researchers make mistakes.

Although there's a sense in which almost any threat to validity can be characterised as a confounder or an artefact, they're pretty vague concepts. So let's have a look at some of the most common examples.

### 歷史效應

**History effects** refer to the possibility that specific events may occur during the study that might influence the outcome measure. For instance, something might happen in between a pretest and a post-test. Or in-between testing participant 23 and participant 24. Alternatively, it might be that you're looking at a paper from an older study that was perfectly valid for its time, but the world has changed enough since then that the conclusions are no longer trustworthy. Examples of things that would count as history effects are:

-   You're interested in how people think about risk and uncertainty. You started your data collection in December 2010. But finding participants and collecting data takes time, so you're still finding new people in February 2011. Unfortunately for you (and even more unfortunately for others), the Queensland floods occurred in January 2011 causing billions of dollars of damage and killing many people. Not surprisingly, the people tested in February 2011 express quite different beliefs about handling risk than the people tested in December 2010. Which (if any) of these reflects the "true" beliefs of participants? I think the answer is probably both. The Queensland floods genuinely changed the beliefs of the Australian public, though possibly only temporarily. The key thing here is that the "history" of the people tested in February is quite different to people tested in December.

-   You're testing the psychological effects of a new anti-anxiety drug. So what you do is measure anxiety before administering the drug (e.g., by self-report, and taking physiological measures). Then you administer the drug, and afterwards you take the same measures. In the middle however, because your lab is in Los Angeles, there's an earthquake which increases the anxiety of the participants.

### 發展成熟效應

As with history effects, **maturational effects** are fundamentally about change over time. However, maturation effects aren't in response to specific events. Rather, they relate to how people change on their own over time. We get older, we get tired, we get bored, etc. Some examples of maturation effects are:

-   When doing developmental psychology research you need to be aware that children grow up quite rapidly. So, suppose that you want to find out whether some educational trick helps with vocabulary size among 3 year olds. One thing that you need to be aware of is that the vocabulary size of children that age is growing at an incredible rate (multiple words per day) all on its own. If you design your study without taking this maturational effect into account, then you won't be able to tell if your educational trick works.
-   When running a very long experiment in the lab (say, something that goes for 3 hours) it's very likely that people will begin to get bored and tired, and that this maturational effect will cause performance to decline regardless of anything else going on in the experiment

### 重覆測試效應

An important type of history effect is the effect of **repeated testing**. Suppose I want to take two measurements of some psychological construct (e.g., anxiety). One thing I might be worried about is if the first measurement has an effect on the second measurement. In other words, this is a history effect in which the "event" that influences the second measurement is the first measurement itself! This is not at all uncommon. Examples of this include:

-   Learning and practice: e.g., "intelligence" at time 2 might appear to go up relative to time 1 because participants learned the general rules of how to solve "intelligence-test-style" questions during the first testing session.
-   Familiarity with the testing situation: e.g., if people are nervous at time 1, this might make performance go down. But after sitting through the first testing situation they might calm down a lot precisely because they've seen what the testing looks like.
-   Auxiliary changes caused by testing: e.g., if a questionnaire assessing mood is boring then mood rating at measurement time 2 is more likely to be "bored" precisely because of the boring measurement made at time 1.

### 選擇偏誤

**Selection bias** is a pretty broad term. Suppose that you're running an experiment with two groups of participants where each group gets a different "treatment", and you want to see if the different treatments lead to different outcomes. However, suppose that, despite your best efforts, you've ended up with a gender imbalance across groups (say, group A has 80% females and group B has 50% females). It might sound like this could never happen but, trust me, it can. This is an example of a selection bias, in which the people "selected into" the two groups have different characteristics. If any of those characteristics turns out to be relevant (say, your treatment works better on females than males) then you're in a lot of trouble.

### 個體特質差異

When thinking about the effects of attrition, it is sometimes helpful to distinguish between two different types. The first is **homogeneous attrition**, in which the attrition effect is the same for all groups, treatments or conditions. In the example I gave above, the attrition would be homogeneous if (and only if) the easily bored participants are dropping out of all of the conditions in my experiment at about the same rate. In general, the main effect of homogeneous attrition is likely to be that it makes your sample unrepresentative. As such, the biggest worry that you'll have is that the generalisability of the results decreases. In other words, you lose external validity.

The second type of attrition is **heterogeneous attrition**, in which the attrition effect is different for different groups. More often called **differential attrition**, this is a kind of selection bias that is caused by the study itself. Suppose that, for the first time ever in the history of psychology, I manage to find the perfectly balanced and representative sample of people. I start running "Dani's incredibly long and tedious experiment" on my perfect sample but then, because my study is incredibly long and tedious, lots of people start dropping out. I can't stop this. Participants absolutely have the right to stop doing any experiment, any time, for whatever reason they feel like, and as researchers we are morally (and professionally) obliged to remind people that they do have this right. So, suppose that "Dani's incredibly long and tedious experiment" has a very high drop out rate. What do you suppose the odds are that this drop out is random? Answer: zero. Almost certainly the people who remain are more conscientious, more tolerant of boredom, etc., than those that leave. To the extent that (say) conscientiousness is relevant to the psychological phenomenon that I care about, this attrition can decrease the validity of my results.

Here's another example. Suppose I design my experiment with two conditions. In the "treatment" condition, the experimenter insults the participant and then gives them a questionnaire designed to measure obedience. In the "control" condition, the experimenter engages in a bit of pointless chitchat and then gives them the questionnaire. Leaving aside the questionable scientific merits and dubious ethics of such a study, let's have a think about what might go wrong here. As a general rule, when someone insults me to my face I tend to get much less co-operative. So, there's a pretty good chance that a lot more people are going to drop out of the treatment condition than the control condition. And this drop out isn't going to be random. The people most likely to drop out would probably be the people who don't care all that much about the importance of obediently sitting through the experiment. Since the most bloody minded and disobedient people all left the treatment group but not the control group, we've introduced a confound: the people who actually took the questionnaire in the treatment group were already more likely to be dutiful and obedient than the people in the control group. In short, in this study insulting people doesn't make them more obedient. It makes the more disobedient people leave the experiment! The internal validity of this experiment is completely shot.

### 無回應偏誤

**Non-response bias** is closely related to selection bias and to differential attrition. The simplest version of the problem goes like this. You mail out a survey to 1000 people but only 300 of them reply. The 300 people who replied are almost certainly not a random subsample. People who respond to surveys are systematically different to people who don't. This introduces a problem when trying to generalise from those 300 people who replied to the population at large, since you now have a very non-random sample. The issue of non-response bias is more general than this, though. Among the (say) 300 people that did respond to the survey, you might find that not everyone answers every question. If (say) 80 people chose not to answer one of your questions, does this introduce problems? As always, the answer is maybe. If the question that wasn't answered was on the last page of the questionnaire, and those 80 surveys were returned with the last page missing, there's a good chance that the missing data isn't a big deal; probably the pages just fell off. However, if the question that 80 people didn't answer was the most confrontational or invasive personal question in the questionnaire, then almost certainly you've got a problem. In essence, what you're dealing with here is what's called the problem of **missing data**. If the data that is missing was "lost" randomly, then it's not a big problem. If it's missing systematically, then it can be a big problem.

### 趨向平均數的迴歸

**Regression to the mean** refers to any situation where you select data based on an extreme value on some measure. Because the variable has natural variation it almost certainly means that when you take a subsequent measurement the later measurement will be less extreme than the first one, purely by chance.

Here's an example. Suppose I'm interested in whether a psychology education has an adverse effect on very smart kids. To do this, I find the 20 psychology I students with the best high school grades and look at how well they're doing at university. It turns out that they're doing a lot better than average, but they're not topping the class at university even though they did top their classes at high school. What's going on? The natural first thought is that this must mean that the psychology classes must be having an adverse effect on those students. However, while that might very well be the explanation, it's more likely that what you're seeing is an example of "regression to the mean". To see how it works, let's take a moment to think about what is required to get the best mark in a class, regardless of whether that class be at high school or at university. When you've got a big class there are going to be lots of very smart people enrolled. To get the best mark you have to be very smart, work very hard, and be a bit lucky. The exam has to ask just the right questions for your idiosyncratic skills, and you have to avoid making any dumb mistakes (we all do that sometimes) when answering them. And that's the thing, whilst intelligence and hard work are transferable from one class to the next, luck isn't. The people who got lucky in high school won't be the same as the people who get lucky at university. That's the very definition of "luck". The consequence of this is that when you select people at the very extreme values of one measurement (the top 20 students), you're selecting for hard work, skill and luck. But because the luck doesn't transfer to the second measurement (only the skill and work), these people will all be expected to drop a little bit when you measure them a second time (at university). So their scores fall back a little bit, back towards everyone else. This is regression to the mean.

Regression to the mean is surprisingly common. For instance, if two very tall people have kids their children will tend to be taller than average but not as tall as the parents. The reverse happens with very short parents. Two very short parents will tend to have short children, but nevertheless those kids will tend to be taller than the parents. It can also be extremely subtle. For instance, there have been studies done that suggested that people learn better from negative feedback than from positive feedback. However, the way that people tried to show this was to give people positive reinforcement whenever they did good, and negative reinforcement when they did bad. And what you see is that after the positive reinforcement people tended to do worse, but after the negative reinforcement they tended to do better. But notice that there's a selection bias here! When people do very well, you're selecting for "high" values, and so you should expect, because of regression to the mean, that performance on the next trial should be worse regardless of whether reinforcement is given. Similarly, after a bad trial, people will tend to improve all on their own. The apparent superiority of negative feedback is an artefact caused by regression to the mean (see @Kahneman1973, for discussion).

### 實驗者偏誤

**Experimenter bias** can come in multiple forms. The basic idea is that the experimenter, despite the best of intentions, can accidentally end up influencing the results of the experiment by subtly communicating the "right answer" or the "desired behaviour" to the participants. Typically, this occurs because the experimenter has special knowledge that the participant does not, for example the right answer to the questions being asked or knowledge of the expected pattern of performance for the condition that the participant is in. The classic example of this happening is the case study of "Clever Hans", which dates back to 1907 [@Pfungst1911]. Clever Hans was a horse that apparently was able to read and count and perform other human like feats of intelligence. After Clever Hans became famous, psychologists started examining his behaviour more closely. It turned out that, not surprisingly, Hans didn't know how to do maths. Rather, Hans was responding to the human observers around him, because the humans did know how to count and the horse had learned to change its behaviour when people changed theirs.

The general solution to the problem of experimenter bias is to engage in double blind studies, where neither the experimenter nor the participant knows which condition the participant is in or knows what the desired behaviour is. This provides a very good solution to the problem, but it's important to recognise that it's not quite ideal, and hard to pull off perfectly. For instance, the obvious way that I could try to construct a double blind study is to have one of my Ph.D. students (one who doesn't know anything about the experiment) run the study. That feels like it should be enough. The only person (me) who knows all the details (e.g., correct answers to the questions, assignments of participants to conditions) has no interaction with the participants, and the person who does all the talking to people (the Ph.D. student) doesn't know anything. Except for the reality that the last part is very unlikely to be true. In order for the Ph.D. student to run the study effectively they need to have been briefed by me, the researcher. And, as it happens, the Ph.D. student also knows me and knows a bit about my general beliefs about people and psychology (e.g., I tend to think humans are much smarter than psychologists give them credit for). As a result of all this, it's almost impossible for the experimenter to avoid knowing a little bit about what expectations I have. And even a little bit of knowledge can have an effect. Suppose the experimenter accidentally conveys the fact that the participants are expected to do well in this task. Well, there's a thing called the "Pygmalion effect", where if you expect great things of people they'll tend to rise to the occasion. But if you expect them to fail then they'll do that too. In other words, the expectations become a self-fulfilling prophesy.

### 需求效應與反應性

When talking about experimenter bias, the worry is that the experimenter's knowledge or desires for the experiment are communicated to the participants, and that these can change people's behaviour [@Rosenthal1966]. However, even if you manage to stop this from happening, it's almost impossible to stop people from knowing that they're part of a psychological study. And the mere fact of knowing that someone is watching or studying you can have a pretty big effect on behaviour. This is generally referred to as **reactivity** or **demand effects**. The basic idea is captured by the Hawthorne effect: people alter their performance because of the attention that the study focuses on them. The effect takes its name from a study that took place in the "Hawthorne Works" factory outside of Chicago (see @Adair1984). This study, from the 1920s, looked at the effects of factory lighting on worker productivity. But, importantly, change in worker behaviour occurred because the workers knew they were being studied, rather than any effect of factory lighting.

To get a bit more specific about some of the ways in which the mere fact of being in a study can change how people behave, it helps to think like a social psychologist and look at some of the roles that people might adopt during an experiment but might not adopt if the corresponding events were occurring in the real world:

-   The *good participant* tries to be too helpful to the researcher. He or she seeks to figure out the experimenter's hypotheses and confirm them.
-   The *negative participant* does the exact opposite of the good participant. He or she seeks to break or destroy the study or the hypothesis in some way.
-   The *faithful participant* is unnaturally obedient. He or she seeks to follow instructions perfectly, regardless of what might have happened in a more realistic setting.
-   The *apprehensive participant* gets nervous about being tested or studied, so much so that his or her behaviour becomes highly unnatural, or overly socially desirable.

### 安慰劑效應

The **placebo effect** is a specific type of demand effect that we worry a lot about. It refers to the situation where the mere fact of being treated causes an improvement in outcomes. The classic example comes from clinical trials. If you give people a completely chemically inert drug and tell them that it's a cure for a disease, they will tend to get better faster than people who aren't treated at all. In other words, it is people's belief that they are being treated that causes the improved outcomes, not the drug.

However, the current consensus in medicine is that true placebo effects are quite rare and most of what was previously considered placebo effect is in fact some combination of natural healing (some people just get better on their own), regression to the mean and other quirks of study design. Of interest to psychology is that the strongest evidence for at least some placebo effect is in self-reported outcomes, most notably in treatment of pain [@hrobjartsson2010].

### 情境、測量、及小群體效應

In some respects, these terms are a catch-all term for "all other threats to external validity". They refer to the fact that the choice of sub-population from which you draw your participants, the location, timing and manner in which you run your study (including who collects the data) and the tools that you use to make your measurements might all be influencing the results. Specifically, the worry is that these things might be influencing the results in such a way that the results won't generalise to a wider array of people, places and measures.

### 詐欺、欺暪與自我欺暪

> *It is difficult to get a man to understand something, when his salary depends on his not understanding it.*\
> - Upton Sinclair

There's one final thing I feel I should mention. While reading what the textbooks often have to say about assessing the validity of a study I couldn't help but notice that they seem to make the assumption that the researcher is honest. I find this hilarious. While the vast majority of scientists are honest, in my experience at least, some are not.[^02-a-brief-introduction-to-research-design-6] Not only that, as I mentioned earlier, scientists are not immune to belief bias. It's easy for a researcher to end up deceiving themselves into believing the wrong thing, and this can lead them to conduct subtly flawed research and then hide those flaws when they write it up. So you need to consider not only the (probably unlikely) possibility of outright fraud, but also the (probably quite common) possibility that the research is unintentionally "slanted". I opened a few standard textbooks and didn't find much of a discussion of this problem, so here's my own attempt to list a few ways in which these issues can arise:

[^02-a-brief-introduction-to-research-design-6]: Some people might argue that if you're not honest then you're not a real scientist. Which does have some truth to it I guess, but that's disingenuous (look up the "No true Scotsman" fallacy). The fact is that there are lots of people who are employed ostensibly as scientists, and whose work has all of the trappings of science, but who are outright fraudulent. Pretending that they don't exist by saying that they're not scientists is just muddled thinking.

-   **資料造假(Data fabrication)**. Sometimes, people just make up the data. This is occasionally done with "good" intentions. For instance, the researcher believes that the fabricated data do reflect the truth, and may actually reflect "slightly cleaned up" versions of actual data. On other occasions, the fraud is deliberate and malicious. Some high-profile examples where data fabrication has been alleged or shown include Cyril Burt (a psychologist who is thought to have fabricated some of his data), Andrew Wakefield (who has been accused of fabricating his data connecting the MMR vaccine to autism) and Hwang Woo-suk (who falsified a lot of his data on stem cell research).
-   **惡作劇(Hoaxes)**. Hoaxes share a lot of similarities with data fabrication, but they differ in the intended purpose. A hoax is often a joke, and many of them are intended to be (eventually) discovered. Often, the point of a hoax is to discredit someone or some field. There's quite a few well known scientific hoaxes that have occurred over the years (e.g., Piltdown man) and some were deliberate attempts to discredit particular fields of research (e.g., the Sokal affair).
-   **資料不實(Data misrepresentation)**. While fraud gets most of the headlines, it's much more common in my experience to see data being misrepresented. When I say this I'm not referring to newspapers getting it wrong (which they do, almost always). I'm referring to the fact that often the data don't actually say what the researchers think they say. My guess is that, almost always, this isn't the result of deliberate dishonesty but instead is due to a lack of sophistication in the data analyses. For instance, think back to the example of Simpson's paradox that I discussed in the beginning of this book. It's very common to see people present "aggregated" data of some kind and sometimes, when you dig deeper and find the raw data yourself you find that the aggregated data tell a different story to the disaggregated data. Alternatively, you might find that some aspect of the data is being hidden, because it tells an inconvenient story (e.g., the researcher might choose not to refer to a particular variable). There's a lot of variants on this, many of which are very hard to detect.
-   **“不良”研究設計(Study "misdesign")**. Okay, this one is subtle. Basically, the issue here is that a researcher designs a study that has built-in flaws and those flaws are never reported in the paper. The data that are reported are completely real and are correctly analysed, but they are produced by a study that is actually quite wrongly put together. The researcher really wants to find a particular effect and so the study is set up in such a way as to make it "easy" to (artefactually) observe that effect. One sneaky way to do this, in case you're feeling like dabbling in a bit of fraud yourself, is to design an experiment in which it's obvious to the participants what they're "supposed" to be doing, and then let reactivity work its magic for you. If you want you can add all the trappings of double blind experimentation but it won't make a difference since the study materials themselves are subtly telling people what you want them to do. When you write up the results the fraud won't be obvious to the reader. What's obvious to the participant when they're in the experimental context isn't always obvious to the person reading the paper. Of course, the way I've described this makes it sound like it's always fraud. Probably there are cases where this is done deliberately, but in my experience the bigger concern has been with unintentional misdesign. The researcher believes and so the study just happens to end up with a built in flaw, and that flaw then magically erases itself when the study is written up for publication.
-   **資料探勘與後設檢定(Data mining & post hoc hypothesising)**. Another way in which the authors of a study can more or less misrepresent the data is by engaging in what's referred to as "data mining" (see Gelman and Loken 2014, for a broader discussion of this as part of the "garden of forking paths" in statistical analysis). As we'll discuss later, if you keep trying to analyse your data in lots of different ways, you'll eventually find something that "looks" like a real effect but isn't. This is referred to as "data mining". It used to be quite rare because data analysis used to take weeks, but now that everyone has very powerful statistical software on their computers it's becoming very common. Data mining per se isn't "wrong", but the more that you do it the bigger the risk you're taking. The thing that is wrong, and I suspect is very common, is unacknowledged data mining. That is, the researcher runs every possible analysis known to humanity, finds the one that works, and then pretends that this was the only analysis that they ever conducted. Worse yet, they often "invent" a hypothesis after looking at the data to cover up the data mining. To be clear. It's not wrong to change your beliefs after looking at the data, and to reanalyse your data using your new "post hoc" hypotheses. What is wrong (and I suspect common) is failing to acknowledge that you've done. If you acknowledge that you did it then other researchers are able to take your behaviour into account. If you don't, then they can't. And that makes your behaviour deceptive. Bad!
-   **發表偏誤與自我審查(Publication bias & self-censoring)**. Finally, a pervasive bias is "non-reporting" of negative results. This is almost impossible to prevent. Journals don't publish every article that is submitted to them. They prefer to publish articles that find "something". So, if 20 people run an experiment looking at whether reading Finnegans Wake causes insanity in humans, and 19 of them find that it doesn't, which one do you think is going to get published? Obviously, it's the one study that did find that Finnegans Wake causes insanity.[^02-a-brief-introduction-to-research-design-7] This is an example of a publication bias. Since no-one ever published the 19 studies that didn't find an effect, a naive reader would never know that they existed. Worse yet, most researchers "internalise" this bias and end up self-censoring their research. Knowing that negative results aren't going to be accepted for publication, they never even try to report them. As a friend of mine says "for every experiment that you get published, you also have 10 failures". And she's right. The catch is, while some (maybe most) of those studies are failures for boring reasons (e.g. you stuffed something up) others might be genuine "null" results that you ought to acknowledge when you write up the "good" experiment. And telling which is which is often hard to do. A good place to start is a paper by @Ioannidis2005 with the depressing title "Why most published research findings are false". I'd also suggest taking a look at work by @Kuhberger2014 presenting statistical evidence that this actually happens in psychology.

[^02-a-brief-introduction-to-research-design-7]: Clearly, the real effect is that only insane people would even try to read *Finnegans Wake*

There's probably a lot more issues like this to think about, but that'll do to start with. What I really want to point out is the blindingly obvious truth that real world science is conducted by actual humans, and only the most gullible of people automatically assumes that everyone else is honest and impartial. Actual scientists aren't usually that naive, but for some reason the world likes to pretend that we are, and the textbooks we usually write seem to reinforce that stereotype.

## 本章小結

This chapter isn't really meant to provide a comprehensive discussion of psychological research methods. It would require another volume just as long as this one to do justice to the topic. However, in real life statistics and study design are so tightly intertwined that it's very handy to discuss some of the key topics. In this chapter, I've briefly discussed the following topics:

-   [Introduction to psychological measurement]. What does it mean to operationalise a theoretical construct? What does it mean to have variables and take measurements?
-   [Scales of measurement] and types of variables. Remember that there are two different distinctions here. There's the difference between discrete and continuous data, and there's the difference between the four different scale types (nominal, ordinal, interval and ratio).
-   [Assessing the reliability of a measurement]. If I measure the "same" thing twice, should I expect to see the same result? Only if my measure is reliable. But what does it mean to talk about doing the "same" thing? Well, that's why we have different types of reliability. Make sure you remember what they are.
-   [The "role" of variables: predictors and outcomes]. What roles do variables play in an analysis? Can you remember the difference between predictors and outcomes? Dependent and independent variables? Etc.
-   [Experimental and non-experimental research] designs. What makes an experiment an experiment? Is it a nice white lab coat, or does it have something to do with researcher control over variables?
-   [Assessing the validity of a study]. Does your study measure what you want it to? How might things go wrong? And is it my imagination, or was that a very long list of possible ways in which things can go wrong?

All this should make clear to you that study design is a critical part of research methodology. I built this chapter from the classic little book by @Campbell1963, but there are of course a large number of textbooks out there on research design. Spend a few minutes with your favourite search engine and you'll find dozens.
